Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-06 agujournaltemplate.tex
FINAL VERSION for checks by coauthors

% MY OWN ADDED PACKAGES TO RENDER UNITS
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{siunitx}

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. 
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

    The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between MEI v.2 and other variables in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month time lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
    The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 anomalies, which represent La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response to the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in cluster 1 (see Figure \ref{fig:clustering} $e$). This is consistent with previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby increasing upwelling in the Cariaco Basin \cite{huang_enso_2023}.

    The in-situ measurements that best predicted changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order. Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For diatoms, a strong threshold response was observed above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drove blooms, resulting in predictable changes in cell counts. Similarly, haptophytes and diatoms exhibited a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.

@@

\appendix

\section{Correlation of data}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
\caption{Correlation matrix with clustering of environmental variables, diversity metrics and cell counts. Correlation was calcualted using a Spearman's rank correlation coefficient, the magnitude of the coefficient is indicated by the color of the circle. The size of the circle and significance level marking represents the calculated significance of the correlation deviating from 0 (p-value: ***\textless0.001, **\textless0.01, *\textless0.05). The time series were aditionally ordered by hierarchical clustering via the complete linkage method into 5 groups indicated by the black square border.}
\label{fig:sup:correlation}
\end{figure}

\section{Gradient Forest analysis supplement}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA2_GF_output_Supplement_v1.pdf}
\caption{Cumulative importance curves for all predictor variables not shown in the main figure. Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups. Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
\label{fig:sup:GFoutput_lags_extra}
\end{figure}

@@

@@

@@

\end{tabular}

@@

\caption{Table \ref{tab:sup:GenusFuncGroups} cont.:}
\begin{tabular}{@{}ll@{}}
\textbf{Dinoflagellates} cont.    &                                                            \\
\textit{Dinophysis}        & Ehrenberg, 1839                                            \\
\textit{Diplopsalis}       & R.S.Bergh, 1881                                            \\
\textit{Glenodinium}       & Ehrenberg, 1836     
\\
\textit{Gonyaulax}         & Diesing, 1866                                              \\
\textit{Gymnodinium}       & F.Stein, 1878                                              \\
\textit{Gyrodinium}        & Kofoid \& Swezy, 1921                                      \\
\textit{Heterocapsa}       & F.Stein, 1883                                              \\
\textit{Lingulodinium}     & D.Wall, 1967                                               \\
\textit{Mesoporos}         & Lillick, 1937                                              \\
\textit{Ornithocercus}     & Stein, 1883                                                \\
\textit{Oxytoxum}          & Stein, 1883                                                \\
\textit{Phalacroma}        & Stein, 1883                                                \\
\textit{Oxyphysis}         & Kofoid, 1926                                               \\
\textit{Podolampas}        & F.Stein, 1883                                              \\
\textit{Pronoctiluca}      & Fabre-Domergue, 1889                                       \\
\textit{Prorocentrum}      & Ehrenberg, 1834                                            \\
\textit{Exuviella}         & Cienkowski, 1881                                           \\
\textit{Protoperidinium}   & Bergh, 1881                                                \\
\textit{Pyrocystis}        & J.Murray ex Haeckel, 1890                                  \\
\textit{Pyrophacus}        & F.Stein, 1883                                              \\
\textit{Scrippsiella}      & Balech ex A.R.Loeblich III, 1965                           \\
\textit{Sphaerodinium}     & Woloszynska, 1916                                          \\
\textit{Triadinium}        & Dodge, 1981                                                \\
\textit{Tripos}            & Bory de Saint-Vincent, 1823                                \\
\textbf{Haptophytes}        &                                                            \\
\textit{Calcidiscus}       & Kamptner, 1950                                             \\
\textit{Calciopappus}      & Gaarder \& Ramsfjell, 1954                                 \\
\textit{Calciosolenia}     & Gran, 1912                                                 \\
\textit{Calyptrosphaera}   & Lohmann, 1902                                              \\
\textit{Coccolithus}       & E.H.L.Schwarz, 1894                                        \\
\textit{Cyclolithella}     & Loeblich \& Tappan, 1963                                   \\
\textit{Emiliania}         & W.W.Hay \& H.P.Mohler, 1967                                \\
\textit{Michaelsarsia}     & Gran, 1912                                                 \\
\textit{Ophiaster}         & Gran, 1912                                                 \\
\textit{Syracosphaera}     & Lohmann, 1902                                              \\
\textit{Umbellosphaera}    & Paasche, 1955                                              \\
\textit{Umbilicosphaera}   & Lohmann, 1902                                              \\
\textbf{Others}            &                                                            \\
\textit{Dictyocha}         & Ehrenberg, 1837                                            \\
\textit{Eutreptia}         & Perty, 1852                                                \\
\textit{Eutreptiella}      & A.M.Cunha, 1913                                            \\
\textit{Mesodinium}        & von Stein                                                 
\end{tabular}
\end{table}

@@

%!
\acknowledgments
We thank Digna Rueda-Roa, Claudia Benitez-Nelson, and James Pinckney for their valuable input during the early stages of this research.

>>>>

% MY OWN ADDED PACKAGES TO RENDER UNITS
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[section]{placeins}

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.

    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

    The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During El Niño (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. This would imply a general reduction in upwelling strength during El Niño events in the Cariaco Basin, which has indeed been observed during the period from 1997 to 1998 \cite{romero_seasonal_2009}. The MEI v.2 is significantly positively correlated to local evaporation and salinity and negatively correlated to the AMO index, however it is not significantly correlated to local upwelling-related variables (see Figure \ref{fig:sup:correlation} in Appendix). The link here is less clear; while generally trade wind strength and SST are anti-correlated in the Caribbean Sea, positive trade wind anomalies have been observed during during El Niño conditions and the reverse during La Niña \cite{chang_coupled_2013}. \citeA{taylor_ecosystem_2012} found no strong correlation between MEI v.2 and other variables in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month time lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
    The threshold responses indicate a shift in the abundances of dinoflagellates for negative MEI v.2 anomalies, which represent La Niña conditions. Findings from Bayesian modeling showed that dinoflagellates responding positively to warmer, stratified conditions in the Cariaco Basin \cite{mutshinda_environmental_2013}. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response, but show a slightly stronger response in the positive range of the MEI v.2. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in the high upwelling regime of cluster 1 (see Figure \ref{fig:clustering} $e$). This could be explained by previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby potentially increasing upwelling strength in the Cariaco Basin \cite{huang_enso_2023}.

    The in-situ measurements that best predicted changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order. Haptophytes and diatoms exhibited a steep threshold around \qtyrange{22}{24}{\celsius}, indicating the strong effect on cell counts at low surface temperatures that are indicative of strong upwelling. Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For diatoms, a strong threshold response was observed above \qty{7}{\micro M}, indicating that intense upwelling conditions drove blooms, resulting in predictable changes in cell counts.

@@

\appendix
\newpage
\section{Correlation of data}

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
\caption{Correlation matrix with clustering of environmental variables, diversity metrics and cell counts. Correlation was calcualted using a Spearman's rank correlation coefficient, the magnitude of the coefficient is indicated by the color of the circle. The size of the circle and significance level marking represents the calculated significance of the correlation deviating from 0 (p-value: ***\textless0.001, **\textless0.01, *\textless0.05). The time series were aditionally ordered by hierarchical clustering via the complete linkage method into 5 groups indicated by the black square border.}
\label{fig:sup:correlation}
\end{figure}

\newpage
\section{Gradient Forest analysis supplement}

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{fig/FigureA2_GF_output_Supplement_v1.pdf}
\caption{Cumulative importance curves for all predictor variables not shown in the main figure. Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups. Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
\label{fig:sup:GFoutput_lags_extra}
\end{figure}

@@

%\end{tabular}}

%\label{tab:sup:GenusFuncGroups}
%\end{table}

%\begin{table}

%\caption{Table \ref{tab:sup:GenusFuncGroups} cont.:}
%\begin{tabular}{@{}ll@{}}

@@

%\resizebox*{\textheight}{!}{
\resizebox*{!}{\dimexpr\textheight-\lineskip\relax}{%

@@

\textit{Dinophysis}        & Ehrenberg, 1839                                            \\
\textit{Diplopsalis}       & R.S.Bergh, 1881                                            \\
\textit{Glenodinium}       & Ehrenberg, 1836     
\\
\textit{Gonyaulax}         & Diesing, 1866                                              \\
\textit{Gymnodinium}       & F.Stein, 1878                                              \\
\textit{Gyrodinium}        & Kofoid \& Swezy, 1921                                      \\
\textit{Heterocapsa}       & F.Stein, 1883                                              \\
\textit{Lingulodinium}     & D.Wall, 1967                                               \\
\textit{Mesoporos}         & Lillick, 1937                                              \\
\textit{Ornithocercus}     & Stein, 1883                                                \\
\textit{Oxytoxum}          & Stein, 1883                                                \\
\textit{Phalacroma}        & Stein, 1883                                                \\
\textit{Oxyphysis}         & Kofoid, 1926                                               \\
\textit{Podolampas}        & F.Stein, 1883                                              \\
\textit{Pronoctiluca}      & Fabre-Domergue, 1889                                       \\
\textit{Prorocentrum}      & Ehrenberg, 1834                                            \\
\textit{Exuviella}         & Cienkowski, 1881                                           \\
\textit{Protoperidinium}   & Bergh, 1881                                                \\
\textit{Pyrocystis}        & J.Murray ex Haeckel, 1890                                  \\
\textit{Pyrophacus}        & F.Stein, 1883                                              \\
\textit{Scrippsiella}      & Balech ex A.R.Loeblich III, 1965                           \\
\textit{Sphaerodinium}     & Woloszynska, 1916                                          \\
\textit{Triadinium}        & Dodge, 1981                                                \\
\textit{Tripos}            & Bory de Saint-Vincent, 1823                                \\
\textbf{Haptophytes}        &                                                            \\
\textit{Calcidiscus}       & Kamptner, 1950                                             \\
\textit{Calciopappus}      & Gaarder \& Ramsfjell, 1954                                 \\
\textit{Calciosolenia}     & Gran, 1912                                                 \\
\textit{Calyptrosphaera}   & Lohmann, 1902                                              \\
\textit{Coccolithus}       & E.H.L.Schwarz, 1894                                        \\
\textit{Cyclolithella}     & Loeblich \& Tappan, 1963                                   \\
\textit{Emiliania}         & W.W.Hay \& H.P.Mohler, 1967                                \\
\textit{Michaelsarsia}     & Gran, 1912                                                 \\
\textit{Ophiaster}         & Gran, 1912                                                 \\
\textit{Syracosphaera}     & Lohmann, 1902                                              \\
\textit{Umbellosphaera}    & Paasche, 1955                                              \\
\textit{Umbilicosphaera}   & Lohmann, 1902                                              \\
\textbf{Others}            &                                                            \\
\textit{Dictyocha}         & Ehrenberg, 1837                                            \\
\textit{Eutreptia}         & Perty, 1852                                                \\
\textit{Eutreptiella}      & A.M.Cunha, 1913                                            \\
\textit{Mesodinium}        & von Stein                                                 
\end{tabular}}

@@

@@

%!
\acknowledgments
We thank Digna Rueda-Roa, Claudia Benitez-Nelson, and James Pinckney for their valuable input during the very early stages of this research. Additionally we want to thank all researchers and staff involved in the CARIACO Time Series for the considerable effort of collecting the time series data and providing it openly to the scientific community.

AI tools (specifically Open AI's ChatGPT) were used to aid in spelling and grammar corrections in writing this study.
